---
title: "Functional Programming"
format:
  revealjs:
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: chalkboard.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
engine: knitr
knitr:
  opts_chunk: 
    echo: true
    eval: false
    comment: "#>" 


revealjs-plugins:
  - verticator
  
filters: 
  - reveal-header
  - reveal-auto-agenda
  - code-fullscreen
  - webr

webr: 
  show-startup-message: true
---

# Nested `for` Loops

## Nested `for` Loops

Nested `for` loops are `for` loops within another `for` loop. You can stack these loops as much as needed. Just make sure the index is different for each loop. The general format for a loop goes as follow:

```{r}
#| eval: false
for (i in vector_1){
  for (ii in vector_2){
    perform task
  }
}
```

## Example

Without using the `sd()` function, compute the standard deviation for each column of the matrix:

```{r}
x <- matrix(rnorm(1000), nrow = 10)
```

$$
s^2 = \frac{1}{n-1}\sum^n_{i=1}(x_i-\bar x)^2
$$


# `while` Loops

## `while` Loops

A while loop is a combination of a `for` loop and a `break` statement. The loop will continue indefinitely until a condition becomes false.

```{r}
#| eval: false

# Initial Condition
condition <- starting TRUE condition

while (condition){
  perform task
  condition <- update condition
}
```

## Example

Simulate from a $N(0,1)$ distribution until you have 50 positive numbers.

```{r}
#| code-fold: true
n <- 0
pos_num <- c()

while (n < 51){
  x <- rnorm(1)
  if (x > 0) {
    pos_num <- c(pos_num, x)
    n <- n + 1
  }
}

pos_num
```

## Example

Find the value of $x$ where the function $y=1/x$ relative converges ($\frac{|y_{old}-y_{new}|}{y_{old}}$) at a level of $10^-6$ as $x\rightarrow \infty$.

```{r}
#| code-fold: true

diff <- 10
x <- 2
y_old <- 1
while (diff > 1e-6){
  y_new <- 1 / x
  diff <- abs(y_old - y_new) / y_old
  x <- x + 1
  y_old <- y_new
}


```

## Example

Simulate from a $Binom(1,.2)$ distribution until the sum of the random variables generated is 50.

```{r}
#| code-fold: true

sum_bin <- 0
x <- c()
while (sum_bin <51) {
  x <- c(x, rbinom(1, 1, 0.2))
  sum_bin <- sum(x)
}
sum_bin
length(x)
```


# More Examples

## Example


## Example

Using the code below:

```{r}
x <- rnorm(5000)
```

Create a new vector containing all the positive values of `x`. The new vector should be less than 5000.

Answer:

```{r}
#| code-fold: true

nn <- length(x)
pos <- c()
for (i in 1:nn){
  if (x[i] > 0) {
    pos <- c(pos, x[i])
  }
}
pos
```


## Example

$$
f(x,y) = x^2 + y^2 + \ln(x+y)
$$

Find all the values of $f(x,y)$ for every combination of $x \in \{1, 8, 13, 25, 42, 67, 95\}$ and $y \in \{6, 12, 18, 52, 61, 79, 83\}$

Store values in a $7\times 7$ matrix.

Answer:

```{r}
#| code-fold: true
x <- c(1, 8, 13, 25, 42, 67, 95)
y <- c(6, 12, 18, 52, 61, 79, 83)
res <- matrix(nrow = 7, ncol = 7)
colnames(res) <- as.character(x)
rownames(res) <- as.character(y)

for (i in 1:7){
  for (ii in 1:7){
    res[ii,i] <- x[i]^2 + y[ii]^2 + log(x[i] + y[ii])
  }
}
print(res)

```

# Vectorized Code

## Vectorized code

# Built-in Functions

## Built-in Functions

There are several available functions in R to conduct specific statistical methods or tasks

## Help Documentation

| Section     | Description                                                 |
|:-------------------|:---------------------------------------------------|
| Description | Provides a brief introduction of the function               |
| Usage       | Provides potential usage of the function                    |
| Arguments   | Arguments that the function can take                        |
| Details     | An in depth description of the function                     |
| Value       | Provides information of the output produced by the function |
| Notes       | Any need to know information about the function             |
| Authors     | Developers of the function                                  |
| References  | References to the model and function                        |
| See Also    | Provide information of supporting functions                 |
| Examples    | Examples of the function                                    |

## Generic Functions

Several R objects have a known class attached to it. A specialized object designed to be read by generic functions, such as `summary()` and `plot()`.

For example, the `summary()` is a generic for several types of functions: `summary.aov()`, `summary.lm()`, `summary.glm()`, and many more.

## Commonly-used Function

| Functions  | Description                 |
|:-----------|:----------------------------|
| `aov()`    | Fits an ANOVA Model         |
| `lm()`     | Fits a linear model         |
| `glm()`    | Fits a general linear model |
| `t.test()` | Conducts a t-test           |

# User-built functions

## User-built functions

-   Functions created by the user for analysis

-   Needs to be ran once to the R environment

-   Will be lost when R session is closed

## Anatomy

```{r}
#| eval: false
#| code-line-numbers: "|1-3|4-5|6|7"

name_of_function <- function(data_1, data_2 = NULL, 
                             argument_1, argument_2 = TRUE, argument_3 = NULL,
                             ...){
  # Conduct Task
  # Conduct Task
  output_object <- Tasks
  return(output_object)
}

```

::: fragment
-   `function`: used to construct the function

-   `data1`: first data argument that needs to supplied

-   `data2`: second data argument that does not need to be supplied

-   `argument1`: first argument must be supplied to alter function

-   `argument2`: second argument to alter function, set to `TRUE`

-   `argument3`: third argument that does not need to be supplied

-   `…`: additional arguments supplied to other functions
:::

## Example

Create a function for

$$
y = \ln(x^2)
$$

## Example

Create a function for

$$
f(x) = \left\{\begin{array}{cc}
x^3 & x<0\\
x^2 + 5 & \mathrm{otherwise}
\end{array} \right.
$$

## Example

Create a function for

$$
f(x,y) = \left\{\begin{array}{cc}
x^3 e^y &  x<0\ \\
x^2 + 5 + \ln(y) & \mathrm{otherwise}
\end{array} \right.
$$

## Example

Create the function that allows your to compute the z-score of a specific value `x` using the sampling distribution from a set of data (`y` vector):

$$
z =  \frac{x-\bar y}{\sqrt{s^2_{y}/n_y}}
$$


# `apply()`

## `apply()`

The `apply` function returns a vector, array, or list of values by applying a function to the margins of an array. You will need to specify the following arguments:

-   `X`: an array to be indexed and applied

-   `MARGIN`: specifyng which index(es) to subset by

-   `FUN`: function to be applied

-   `…`: further arguments to be applied to `FUN`, must be labeled

```{r}
apply(X, MARGIN, FUN, ...)
```

## Example

Find the standard deviation of all the columns of the following matrix:

```{r}
x <- matrix(rnorm(1000), nrow = 10)
```

## Example

Find the $25th$, $50th$, and $75th$ quartiles for each row of the following matrix:

```{r}
x <- matrix(rnorm(1000), nrow = 20)
```

# `lapply()`

## `lapply()`

The `lapply` function applies a function to all the elements of a vector or matrix, and it will return a list. You will need to specify the following arguments:

-   `X`: object to be iterated

-   `FUN`: a function to be applied

-   `…`: further arguments to be passed along to `FUN`

```{r}
lapply(X, FUN, ...)
```

## Example

Create a function that returns a labeled list for with the following values: mean, standard deviation, median, max, and min.

# `sapply()`

## `sapply()`

The `sapply()` function will apply a function to each element of a list or vector, and it will return a simplified object, vector, matrix, or array. The `sapply()` function uses 4 main arguments:

-   `X`: a vector or list to be iterated

-   `FUN`: a function to be applied

-   `…`: arguments passed along to `FUN`, must be labeled

-   `simplify`: indicates how to simplify the function, defaults to n-dimensional array based on output

```{r}
sapply(X, FUN, ..., simplify = TRUE)

```

## Example

Using the vector below, compute the length of each string using `sapply` and `str_length()` from `tidyverse`

```{r}
fruits <- c("apple", "banana", "cherry", "date", "elderberry", "fig", "grape", "honeydew", "kiwi", "lemon")

```

## Example

Using the list generated below, compute the mean of each element of the list using `sapply`.

```{r}
# Generate a list of 10 lists, each containing 5 random numbers
lists <- lapply(1:10, function(i) {
  means <- rpois(1, 3)
  rnorm(5, means)
})
```

## Example

Using the vector below, use the `sapply()` to find $\log(x)$ for each value and return a matrix:

```{r}
numbers <- 4:400
```
