---
title: "Simulation Studies"
subtitle: "Zero Models"
format:
  revealjs:
    include-in-header: "math_commands.html"
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: chalkboard.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
engine: knitr
knitr:
  opts_chunk: 
    code-fold: show
    echo: true
    eval: false
    comment: "#>" 
filters: 
  - reveal-header
  - reveal-auto-agenda
  - code-fullscreen
  - webr
webr: 
  show-startup-message: true
  packages: ['ggplot2', 'dplyr', 'stringr']
editor_options: 
  chunk_output_type: console
---

## R Packages

```{r}
#| code-fold: false
#| eval: true
library(extraDistr)
library(mvtnorm)
library(splines2)
library(tidyverse)
library(brms)
theme_set(theme_bw())
```


# Zero-Inflated Models

## Zero-Inflated Models

Zero-inflated models is a mixture model where random variables are generated from 2 or more distributions. 

::: fragment
On distribution that states the value came from 2 possible distributions. All other values could have come from a common distribution. 
:::

::: fragment

:::


## Mixture Models

$$
f(x) = \pi g(x) + (1-\pi)h(x)
$$

-   $\pi\in(0,1)$

## Zero-Inflated Distribution

A zero-inflated distribution is a mixture model, where the observed value of 0 can com from 2 distributions.

::: fragment
This forces the data to produce more 0's than from a standard distribution.
:::


## Zero-Inflated

$$
P(x=0) = \left\{\begin{array}{cc}
\pi + (1-\pi)g(x) & X=0 \\
(1-\pi)g(x) & otherwise
\end{array}
\right.
$$

## Modelling 0

When modeling a ZI-model, we need to determine where the 0 came from.

::: fragment
This is modeled using a Binary (logistic) GLM.
:::

## ZI-Poisson

If you believe that the data comes from a Poisson distribution, but there is an abnormal amount of 0's, It may indicate that 0 may come from 2 sources of distributions.

::: fragment
Hence the Zero-inflated Poison model may be appropriate.
:::

## Simulating ZI-Poisson

### Zero

$$
g(\pi) = -1.85 + 1.1 X_1
$$

### Poisson

$$
h(\lambda) = 4-2X_2
$$

## Simulating ZI-Poisson

```{r}
n <- 10000
x1 <- rnorm(n, mean = 2)
x2 <- rnorm(n, mean = 2.5)
eta <- boot::inv.logit(-1.85 + 1.1 * x1)
zi <- rbinom(n, size = 1, prob = eta)
mu <- exp(4 - 2 * x2)
p <- rpois(n, mu)
y <- (zi == 0) * p
```

## Bayesian Model

```{r}
brm(bf(y ~ x2, 
       zi ~ x1), 
    data = tibble(y, x1, x2),
    family = zero_inflated_poisson(),
    cores = 4)
```


## ZI-Negative Binomial

If you believe that the data comes from a negative binomial distribution, but there is an abnormal amount of 0's, It may indicate that 0 may come from 2 sources of distributions.

::: fragment
Hence the Zero-inflated Poison model may be appropriate.
:::

## Simulating ZI-Negative Binomial

### Zero

$$
g(\pi) = -1.85 + 1.1 X_1
$$

### Negative Binomial

$$
h(\mu) = -0.85-1.3X_2
$$

## Simulating ZI-Negative Binomial

```{r}
n <- 10000
x1 <- rnorm(n, mean = 2)
x2 <- rnorm(n)
eta <- boot::inv.logit(-1.85 + 1.1 * x1)
zi <- rbinom(n, size = 1, prob = eta)
mu <- exp(-0.85 + 1.3 * x2)
p <- rnbinom(n, mu = mu, size = 0.5)
y <- (zi == 0) * p
```

## Bayesian Model

```{r}
brm(bf(y ~ x2, 
       zi ~ x1), 
    data = tibble(y, x1, x2),
    family = zero_inflated_negbinomial(),
    cores = 4)

```


# Hurdle Models

## Hurdle Models

Hurdle models are used to model data where the value 0 is believed to come from one distribution, and all the other values come from a separate distribution.

::: fragment
It can be thought of as data points had to overcome a hurdle to escape 0 to obtain a different value.
:::

## Hurdle Models

$$
P(x=0) = \left\{\begin{array}{cc}
\pi & X=0 \\
(1-\pi)g(x) & otherwise
\end{array}
\right.
$$

## ZI vs Hurdle

:::: {.columns}

::: {.column width="50%"}

### Hurdle Models

$$
\begin{array}{c}
\pi \\
(1-\pi)g(x)
\end{array}
$$
:::

::: {.column width="50%"}
### ZI Models

$$
\begin{array}{cc}
\pi + (1-\pi)g(x) & X=0 \\
(1-\pi)g(x) & otherwise
\end{array}
$$
:::

::::

## Modelling 0

When modeling a ZI-model, we need to determine where the 0 came from.

::: fragment
This is modeled using a Binary (logistic) GLM.
:::

## Hurdle Models

-   Poisson
-   Negative Binomial Model
-   Log-Normal
-   Gamma

## Simulating Hurdle Poisson

### Zero

$$
g(\pi) = 1.1 + 3.1 X
$$

### Poisson

$$
h(\lambda) = 0.75+1.3X
$$


## Hurdle Poisson Model

```{r}
n <- 10000
x <- rnorm(n)
eta <- boot::inv.logit(1.1 + 3.1 * x)
zi <- rbinom(n, 1, eta)
mu <- exp(0.75 + 1.3 * x)
p <- rtpois(n, lambda = mu, a = 0)
y <- (zi == 0) * p

```

## Bayesian Model

```{r}
brm(bf(y ~ x,
       hu ~ x),
    data = tibble(y, x),
    family = hurdle_poisson(),
    cores = 4)
```


## Simulating Hurdle Gamma

### Zero

$$
g(\pi) = 1.1 + 3.1 X
$$

### Poisson

$$
h(\lambda) = 0.75+1.3X
$$

## Hurdle Gamma Model

```{r}
n <- 10000
x <- rnorm(n, mean  = -0.25)
eta <- boot::inv.logit(1.1 + 3.1 * x)
zi <- rbinom(n, 1, eta)
mu <- exp(0.75 + 1.3 * x)
p <- rgamma(n, shape = 2, scale = mu / 2)
y <- (zi == 0) * p
```


## Bayesian model

```{r}
brm(bf(y ~ x, 
       hu ~ x), 
    data = tibble(y, x),
    family = hurdle_gamma(),
    cores = 4)
```


# Cox Proportional Hazard Models

## Cox Proportional Hazard Models

Cox proportional hazard model was used to model the association between a set of predictors and a time-to-event, in the presence of an independent censoring mechanism.

## Data Type

::: incremental
-   Data is typically recorded as time-to-event data

-   For biomedical studies, researchers are interested in time from diagnosis to death, known as time-to-death
:::


## Censoring

::: incremental
-   Censoring is a mechanism where we do not observe the true time-to-event

-   Not all the time is observed

-   Three common types of censoring mechanisms: *Right, Left,* and *Interval*
:::

## Left Censoring

```{r}
#| eval: true
dat <- data.frame(ID = 1:10, 
                  t1 = c(7, 9, 10, 5, 5, 10, 5, 6, 6, 7) , 
                  censored = c(1, 1, 1, 0, 0, 1, 0, 1, 1, 1))
ggplot(dat, aes(x = ID, y = t1, shape = ifelse(censored, "Death", "Censored"))) + geom_point(size = 4) + 
    geom_linerange(aes(ymin = 0, ymax = t1)) +
    geom_hline(yintercept = 5, lty = 2) +
    coord_flip() + 
    scale_shape_manual(name = "Event", values = c(19, 15)) +
    # ggtitle("Left Censoring") + 
    xlab("Patient ID") +  ylab("Months") + 
    theme_bw()

```

## Data Construction

::: incremental
-   $T_i^*$: True time-to-event

-   $C_i$: Censoring Time

-   $T_i=\min(T_i^*,C_i)$: Observed time-to-event

-   $\delta_i = I(T_i^*<C_i)$: Event indicator
:::

## Hazard Function

$$
h(t) = \lim_{\Delta t \rightarrow 0} \frac{P(t \le T^* < t + \Delta t \mid T^* \ge t)}{\Delta t}
$$

## Proportional Hazard Model

$$
h(t \mid \boldsymbol X) = h_0(t) \exp(\boldsymbol\beta^\mathrm T\boldsymbol X)
$$
-   $h_0(t)$: baseline hazard function

-   $\boldsymbol\beta$: regression coefficients

-   $\boldsymbol X$: predictor variables

## Simulated Model

$$
h(t \mid \boldsymbol X) = 1.5t^{0.5} \exp(-1.5X_1 + 2.3X_2)
$$
$$
C_i\sim Exp(3.5)
$$

## Simulating Model

```{r}
n <- 10000
tdf <- tibble()
for (i in 1:10000){
  phi <- 1.5
  x.vec <- c(mvtnorm::rmvnorm(1,c(0,1.5),diag(c(1,.5))))
  invS = function(t, u){
    h = function(s) 
    {
      phi * s^(phi - 1) * exp(-x.vec[1] * 1.5 + x.vec[2] * 2.3)
    }     
    
    integrate(h, lower = 0, upper = t, subdivisions = 2000)$value + log(u)			
  }
  u = runif(1)
  Up <- 1
  Root <- try(uniroot(invS, interval = c(0, Up), u = u)$root, TRUE)
  while(inherits(Root, "try-error") | Root == 0) {
    Up <- Up + 2
    Root <- try(uniroot(invS, interval = c(0, Up), u = u)$root, TRUE)
    if (Root == 0) {u <- runif(1)}
  }
  trueTimes <- if (!inherits(Root, "try-error")) Root else NA
  
  rateC = 3.5
  Ctimes = rexp(1, rate = rateC)
  
  event.times = pmin(trueTimes, Ctimes,1)
  delta_i = as.numeric(trueTimes <= min(Ctimes,1))
  wdf <- tibble(x1 = x.vec[1], x2 = x.vec[2], time = event.times, censor = delta_i)
  tdf <- bind_rows(tdf, wdf)
}

```

## Bayesian Model


```{r}

brm(time | cens(1-censor) ~ x1 + x2,
    data = tdf,
    family = cox(),
    cores = 4)

```

