---
title: "Homework 1"
date: 1/23/24
editor: visual
draft: false
---

**Due 6/14/24 \@ 11:59 PM**

You must comment all your code to receive credit.

Submit the \*.html file to canvas.

For all problems, you must use at least one `for` or `while` loop.

## Problem 1

Using the following code:

```{r}
x <- matrix(rnorm(1500), nrow = 10)
```

Write the code to produce the output if you use the `rowMeans()` on the R object `x`.

## Problem 2

Using the following code:

```{r}
y <- matrix(sample(1:400, 100, replace = T), nrow = 10)
```

Find the median value for each column of the matrix `y`.

## Problem 3

Write the code that will generate the first 1000 numbers of the [Fibonacci Sequence](https://en.wikipedia.org/wiki/Fibonacci_sequence).



## Problem 4

Write a function for the following equation:

$$
f(x, y, z) =\left\{\begin{array}{cc}
x^2+\sqrt y & z  = 0 \\ 
x^2+2x+3+ \log(y) & z = 1
\end{array}\right.
$$

Include any error messages if necessary.

## Problem 5

Create a function that will give you the first `x` Fibonacci numbers. You must use a `for` loop.
